---
title: "Brazil's Students Enrollment"
author: "Bruno Oliveira de Albuquerque"
date: '2022-07-04'
output: 
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
# Defining figure visual
knitr::opts_chunk$set(
 fig.width = 9,
 fig.asp = 0.5,
 fig.align="center",
 out.width = "90%"
)

# Loading libraries
library(tidyverse)
library(janitor)
library(scales)

# Setting working directory to make it easier to manage the directories
setwd('/home/boa50/Desenvolvimento/education-case-study')

# Setting global variables
legend_order <- c('Brazil', 'High income', 'Upper middle income', 
                  'Lower middle income', 'Low income')
color_scale <- 'Set2'

# Function to load the .csv files with cleaned column names, so it will be
# easier to work with the data
read_dataset <- function(file_name, dataset_folder='dataset/') {
  file_path <- paste(dataset_folder, file_name, sep='')
  df <- read_csv(file_path) %>% clean_names()
  return(df)
}
```

## School Enrollment

School enrollment is a good measure to analyze how students are engaged in educational
programs and how they are thinking about their futures as specialized workers.

As children spend more time studying, they could improve their knowledge and be
able to work on a higher demand job in the future.

On this analysis, I try to expose the influence of some indicators on students
enrollment, focused primarily on Brazil's context. The indicators that we will
cover are:

* GDP invested in education
* Children from 7 to 14 years old that are employed
* Income distribution between classes of population

To get the data about students enrollment and the indicators, a dataset form 
World Bank Group was used and more details about it could be obtained from its
[website](https://datacatalog.worldbank.org/search/dataset/0037712/World-Development-Indicators).

I did some cleaning on the dataset so it could be more useful to our analysis.

```{r Preparing the data, message=FALSE}
# Data obtained from https://datacatalog.worldbank.org/search/dataset/0037712/World-Development-Indicators at 04/07/2022
df_countries <- read_dataset('WDICountry.csv')
df_cleaned <- read_dataset('WDIData.csv') %>%
  # Getting data until 2019 to ignore pandemic situation
  select(c(country_code, indicator_code, x2010:x2019)) %>%
  # Removing rows where all year values are null
  filter(rowSums(is.na(.)) < 10) %>%
  # Getting income groups of each country
  merge(read_dataset('WDICountry.csv'), by='country_code') %>%
  select(c(country_code, income_group, indicator_code, x2010:x2019)) %>%
  # Removing countries not associated with any Income Group
  filter(!is.na(income_group)) %>%
  # Creating a different group to Brazil, to make it comparable with other groups
  mutate(income_group=if_else(country_code == 'BRA', 'Brazil', income_group))
```

Below there is a preview on how the dataset become organized.

```{r Showing the cleaned dataset, echo=FALSE}
head(df_cleaned)
```

### How is school enrollment improving over the last years?

To get a perspective of school enrollment, I checked indicators related to
enrollments on main levels of education: primary, secondary and tertiary. The
enrollments were grouped by the Income Group of the countries as defined on
*WDICountry.csv* file. Only Brazil was took off of its group (Upper middle income)
to make it easier to compare with other groups.

The choice of indicators was made by using a spreadsheet with the *WDISeries.csv*
file.

```{r Definining enrollment indicators}
## Primary and secondary indicators were net indicators so they could reflect better the students that needed to be at each level
# SE.PRM.NENR: School enrollment, primary (% net)
# SE.SEC.NENR: School enrollment, secondary (% net)
# SE.TER.ENRR: School enrollment, tertiary (% gross)
enrollment_indicators <- c('SE.PRM.NENR', 'SE.SEC.NENR', 'SE.TER.ENRR')
```

After determining the indicators, we tried to group the data on a way that we 
could show the differences between groups.

```{r Grouping enrollment data}
df_enrollment <- df_cleaned %>%
  filter(indicator_code %in% enrollment_indicators) %>%
  # Getting only the last 5 year to get results based on previous years investments
  select(c(country_code, income_group, indicator_code, x2015:x2019)) %>%
  # pivoting table to make it easier to plot
  pivot_longer(x2015:x2019, names_to='year', values_to='enrollment') %>%
  # cleaning year values to make it easier to use as label
  mutate(year=gsub('x', '', .$year))
```

I also checked the presence of NA values on the metric variable of the dataset.

```{r Checking NA values in enrollments, echo=FALSE}
print(
  paste(
    'Percentage of NA values on enrollment column: ',
    trunc(sum(is.na(df_enrollment$enrollment)) / nrow(df_enrollment)*100), 
    '%', 
    sep=''
    )
  )
```

And as you can see, about 40% of our enrollment variable is filled with NA values.
As this type of values does not change much between years, I used a strategy to
fill the missing values based on previous years and if there isn't a previous year
fill with future data.

```{r Filling missing values in enrollment}
df_enrollment <- df_enrollment %>%
  group_by(country_code, indicator_code) %>%
  # Using the filling strategy to get values from previous years, if it doesn't exist, get from future year
  fill(enrollment, .direction='downup') %>%
  # Filtering to get only the countries with at least 1 not NA value for each indicator
  filter(!is.na(enrollment))
```

To check how each group is improving on their enrollment level, I plotted a line 
chart showing the differences between them.

```{r Students enrollment chart, echo=FALSE, message=FALSE, warning=FALSE}
# setting representative labels to the indicators
indicators_labels <- c(
  'SE.PRM.NENR' = 'Primary Education',
  'SE.SEC.NENR' =  'Secondary Education',
  'SE.TER.ENRR' = 'Tertiary Education'
)

df_enrollment %>%
  group_by(income_group, indicator_code, year) %>%
  summarise(mean=mean(enrollment), sd=sd(enrollment)) %>%
  ggplot(aes(x=year, y=mean, group=income_group)) +
  theme_minimal() +
  geom_line(aes(color=income_group), size=1) +
  geom_ribbon(aes(ymin=mean - sd, ymax=mean + sd, fill=income_group), alpha=.15) +
  expand_limits(y = 0) +
  # changing the order of the legend values
  scale_fill_brewer(breaks=legend_order, palette=color_scale) +
  scale_colour_brewer(breaks=legend_order, palette=color_scale) +
  # formatting the y labels to include a percent sign
  scale_y_continuous(labels=function(y) paste(y, '%', sep='')) +
  facet_wrap(vars(indicator_code), labeller=as_labeller(indicators_labels)) +
  labs(title='Average enrollment per Income Group and Education level', 
       subtitle='The enrollment is calculated based on the age group of the specific education level',
       x='Year', y='Enrollment', color='Income Group', fill='Income Group')
```

As we can see, there is a small difference on the Primary education level and it
increases on the other levels. It could represent a less given importance to the
other levels or some economic/social factor influences this variations. Another
interesting observation is that there is a huge standard deviation (represented
by the color shadows) inside each Income Group.

The problem with this is that children will become less capable of perform on
more specialized jobs, making it difficult to improve the overall education
quality of the countries.

To this analysis, I will give a special attention to secondary education as it
can serve as a baseline to help children acquire great improvements on their
education.

```{r Secondary education enrollment, echo=FALSE, message=FALSE, warning=FALSE}
# Checking the distribution of values about secondary education enrolling for the Upper middle income countries
brazil_value <- df_enrollment %>%
  filter(income_group == 'Brazil' & indicator_code == 'SE.SEC.NENR' & year == '2019')
brazil_value <- brazil_value$enrollment

df_enrollment %>%
  filter(income_group %in% c('Brazil', 'Upper middle income') & 
           indicator_code == 'SE.SEC.NENR' & year == '2019') %>%
  ggplot(aes(x=enrollment)) + 
  theme_minimal() +
  geom_histogram(binwidth=5, fill='#80b1d3') +
  geom_vline(aes(xintercept=brazil_value), color='#757575', size=1, linetype='dashed') +
  geom_text(aes(x=brazil_value+2, y=-.3, label='Brazil')) +
  scale_x_continuous(labels=function(x) paste(x, '%', sep='')) +
  scale_y_continuous(breaks = pretty_breaks()) +
  labs(title='Enrollment in the Secondary education of the Upper middle income group on 2019', 
       x='Enrollment', y='Number of Countries')
```

### TODO

```{r Grouping GDP spent in education data}
# Checking the investment in education based on GDP (https://data.oecd.org/gdp/gross-domestic-product-gdp.htm)
df_gdp_spent_percent <- df_cleaned %>%
  filter(indicator_code  == 'SE.XPD.TOTL.GD.ZS') %>%
  # Pivoting table to make it easier to plot
  pivot_longer(x2010:x2019, names_to='year', values_to='gdp_percent') %>%
  # Cleaning year values to make it easier to use as label
  mutate(year=gsub('x', '', .$year))
```

```{r Checking NA values in GDP spent, echo=FALSE}
print(
  paste(
    'Percentage of NA values on gdp_percent column: ',
    trunc(sum(is.na(df_gdp_spent_percent$gdp_percent)) / nrow(df_gdp_spent_percent)*100), 
    '%', 
    sep=''
  )
)
```

```{r Filling missing values in GDP spent}
df_gdp_spent_percent <- df_gdp_spent_percent %>%
  group_by(country_code, indicator_code) %>%
  fill(gdp_percent, .direction='downup') %>%
  # filtering to get only the countries with at least 1 not NA value for each indicator
  filter(!is.na(gdp_percent))
```

```{r Percentage of GDP spent per group, echo=FALSE, message=FALSE, warning=FALSE}
df_gdp_spent_percent %>%
  group_by(income_group, indicator_code, year) %>%
  summarise(mean=mean(gdp_percent), sd=sd(gdp_percent)) %>%
  ggplot(aes(x=year, y=mean, group=income_group)) +
  theme_minimal() +
  geom_line(aes(color=income_group), size=1) +
  geom_ribbon(aes(ymin=mean - sd, ymax=mean + sd, fill=income_group), alpha=.15) +
  expand_limits(y = 0) +
  # Changing the order of the legend values
  scale_fill_brewer(breaks=legend_order, palette=color_scale) +
  scale_colour_brewer(breaks=legend_order, palette=color_scale) +
  # Formatting the y labels to include a percent sign
  scale_y_continuous(labels=function(y) paste(y, '%', sep='')) +
  labs(title='Average GDP spent in education', 
       x='Year', y='GDP Spent', color='Income Group', fill='Income Group')
```

```{r Grouping GDP per capita spent in education data}
# Getting GDP per capita changes over time
df_gdp <- df_cleaned %>%
  filter(indicator_code  == 'NY.GDP.PCAP.KD') %>%
  # Pivoting table to make it easier to plot
  pivot_longer(x2010:x2019, names_to='year', values_to='gdp_per_capita') %>%
  # Cleaning year values to make it easier to use as label
  mutate(year=gsub('x', '', .$year)) %>%
  group_by(country_code, indicator_code) %>%
  fill(gdp_per_capita, .direction='downup') %>%
  # Filtering to get only the countries with at least 1 not NA value for each indicator
  filter(!is.na(gdp_per_capita))

df_gdp_spent_per_capita <- df_gdp_spent_percent %>%
  # Joining with gdp data to be able to get the gpd expenditure per capita
  merge(df_gdp, by=c('country_code', 'year')) %>%
  rename(income_group=income_group.x) %>%
  select(c(country_code, year, income_group, gdp_percent, gdp_per_capita)) %>%
  # Getting the gdp spent per capita
  mutate(
    gdp_spent_per_capita=(gdp_percent/100) * gdp_per_capita, 
    year=gsub('x', '', year))
```

```{r GDP spent per capita per group, echo=FALSE, message=FALSE, warning=FALSE}
df_gdp_spent_per_capita %>%
  group_by(income_group, year) %>%
  summarise(mean=mean(gdp_spent_per_capita), sd=sd(gdp_spent_per_capita)) %>%
  ggplot(aes(x=year, y=mean, group=income_group)) +
  theme_minimal() +
  geom_line(aes(color=income_group), size=1) +
  geom_ribbon(aes(ymin=mean - sd, ymax=mean + sd, fill=income_group), alpha=.15) +
  expand_limits(y = 0) +
  # changing the order of the legend values
  scale_fill_brewer(breaks=legend_order, palette=color_scale) +
  scale_colour_brewer(breaks=legend_order, palette=color_scale) +
  # formatting the y labels to include a percent sign
  scale_y_continuous(labels=function(y) paste('$', y, sep='')) +
  labs(title='Average GDP per capita spent in education', 
       caption='The values are based on 2015 US$',
       x='Year', y='GDP per capita spent', color='Income Group', fill='Income Group')
```

```{r Distribution of GDP spent per capita of the High income group, echo=FALSE, message=FALSE, warning=FALSE}
# Let's compare how far we are from the majority of high income countries
brazil_value_gdp_per_capita <- df_gdp_spent_per_capita %>%
  filter(income_group == 'Brazil' & year == '2019')
brazil_value_gdp_per_capita <- brazil_value_gdp_per_capita$gdp_spent_per_capita

df_gdp_spent_per_capita %>%
  filter(income_group %in% c('Brazil', 'High income') & year == '2019') %>%
  ggplot(aes(x=gdp_spent_per_capita)) + 
  theme_minimal() +
  geom_histogram(binwidth=200, fill='#80b1d3') +
  geom_vline(aes(xintercept=brazil_value), color='#757575', size=1, linetype='dashed') +
  geom_text(aes(x=brazil_value+200, y=-.3, label='Brazil')) +
  scale_x_continuous(labels=function(x) paste('$', x, sep='')) +
  scale_y_continuous(breaks = pretty_breaks()) +
  labs(title='GDP per capita spent in education of the High income group on 2019 and Brazil', 
       caption='The values are based on 2015 US$',
       x='GDP per capita spent', y='Number of Countries')
```

```{r Filtering data about children employment}
# Getting data on children in employment 7 - 14 year
df_children_employment <- df_cleaned %>% filter(indicator_code  == 'SL.TLF.0714.ZS')
```

```{r Checking NA values in children employment, echo=FALSE}
# Almost 85% of the data is NA
# because of this we will use the strategy to get only the last year from the last five
print(
  paste(
    'Percentage of NA values on years columns: ',
    trunc(sum(rowSums(is.na(df_children_employment))) / (nrow(df_children_employment)*10)*100), 
    '%', 
    sep=''
  )
)
```

```{r Grouping children employment}
df_children_employment <- df_children_employment %>%
  # Getting only the last year from the last five because there is a lot of NA values
  mutate(children_employment=do.call(coalesce, rev(across(x2015:x2019)))) %>%
  select(c(country_code, income_group, children_employment)) %>%
  filter(!is.na(children_employment))
```

```{r Showing children employment per income group, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(df_children_employment, aes(x=income_group, y=children_employment)) + 
  theme_minimal() +
  geom_boxplot(aes(fill=income_group)) +
  # Removing the legend as it is not necessary on this chart
  theme(legend.position='none') +
  # Changing the order of the legend values and setting default colors
  scale_x_discrete(limits=legend_order) +
  scale_y_continuous(labels=function(y) paste(y, '%', sep='')) +
  scale_fill_brewer(palette=color_scale) +
  labs(title='Percentage of children employment per income group', 
       caption='Only children with 7 - 14 years old were considered',
       x='Income Group', y='Children Employment')
```

```{r Definining poverty indicators}
# getting data about poverty !caution
# distribution of income share per group
# SI.DST.05TH.20: Income share held by highest 20%
# SI.DST.04TH.20: Income share held by fourth 20%
# SI.DST.03RD.20: Income share held by third 20%
# SI.DST.02ND.20: Income share held by second 20%
# SI.DST.FRST.20: Income share held by lowest 20%
poverty_indicators <- c('SI.DST.FRST.20', 'SI.DST.02ND.20' , 'SI.DST.03RD.20', 'SI.DST.04TH.20', 'SI.DST.05TH.20')
```

```{r Filtering data about income share}
df_poverty <- df_cleaned %>%
  filter(indicator_code %in% poverty_indicators) %>%
  # Pivoting table to make it easier to plot
  pivot_longer(x2010:x2019, names_to='year', values_to='income_share') %>%
  # Cleaning year values to make it easier to use as label
  mutate(year=gsub('x', '', year))
```

```{r Checking NA values in income share, echo=FALSE}
# Almost 50% of the data has NA values
# because of this we will use the strategy to fill years with previous values or next values
print(
  paste(
    'Percentage of NA values on income_share column: ',
    trunc(sum(is.na(df_poverty$income_share)) / nrow(df_poverty)*100), 
    '%', 
    sep=''
  )
)
```

```{r Grouping income share values, warning=FALSE}
df_poverty <- df_poverty %>%
  group_by(country_code, indicator_code) %>%
  # Filling NA values so that each year could have the same number of countries
  fill(income_share, .direction='downup') %>%
  filter(!is.na(income_share))

df_poverty_grouped <- df_poverty %>%
  group_by(income_group, indicator_code, year) %>%
  summarise(mean=mean(income_share), sd=sd(income_share)) %>%
  # Getting the difference of values between years per group
  mutate(years_difference=abs(mean - lag(mean)))
```

```{r Checking difference in income share values, echo=FALSE, message=FALSE, warning=FALSE}
# As the difference of values between years are so small, we are
# comparing values based only on the last year
print(
  paste(
    'Maximum difference of values between years: ',
    max(df_poverty_grouped$years_difference, na.rm=TRUE), 
    '%', 
    sep=''
  )
)

ggplot(df_poverty_grouped, aes(x=years_difference)) +
  theme_minimal() +
  geom_histogram(binwidth=.025, fill='#80b1d3') +
  scale_x_continuous(labels=function(x) paste(x, '%', sep='')) +
  labs(title='Difference of income share between years', 
       x='Difference between years', y='Number of Values')
```

```{r Show income share variations, echo=FALSE, message=FALSE, warning=FALSE}
# Setting representative labels to the indicators
poverty_indicators_labels <- c(
  'Highest 20%',
  'Fourth 20%',
  'Third 20%',
  'Second 20%',
  'Lowest 20%'
)

df_poverty_grouped %>%
  filter(year == '2019') %>%
  ggplot(aes(
    # Changing the order of the legend values
    fill=factor(indicator_code, levels=rev(poverty_indicators)),
    y=mean, x=income_group)) + 
  theme_minimal() +
  geom_bar(position=position_dodge(), stat='identity') +
  # Inserting the standard deviation bars https://bioinformatics.stackexchange.com/questions/11222/stacked-bargraph-with-error-bars
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.3, position=position_dodge(0.9), color='#757575') +
  # Setting other colors to differentiate from previous charts
  # Got this custom pallet based on https://colors.dopely.top/palette-generator/ToL9uxh2Egw
  scale_fill_manual(values=c('#82AC85','#A9C7AC','#BFD8C4','#E2CFC9','#D1ACA5'), 
                    labels=poverty_indicators_labels) + 
  scale_x_discrete(limits=legend_order) +
  scale_y_continuous(labels=function(y) paste(y, '%', sep='')) +
  labs(title='Percentage of income share per Population Group and Income Group on 2019', 
       x='Income Group', y='Income Share', fill='Population Share Group')
```