---
title: "Brazil's Students Enrollment"
author: "Bruno Oliveira de Albuquerque"
date: '2022-07-04'
output: 
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
# Defining figure visual
knitr::opts_chunk$set(
 fig.width = 9,
 fig.asp = 0.5,
 fig.align="center",
 out.width = "90%"
)

# Loading libraries
library(tidyverse)
library(janitor)
library(scales)

# Setting working directory to make it easier to manage the directories
setwd('/home/boa50/Desenvolvimento/education-case-study')

# Setting global variables
legend_order <- c('Brazil', 'High income', 'Upper middle income', 
                  'Lower middle income', 'Low income')
color_scale <- 'Set2'

# Function to load the .csv files with cleaned column names, so it will be
# easier to work with the data
read_dataset <- function(file_name, dataset_folder='dataset/') {
  file_path <- paste(dataset_folder, file_name, sep='')
  df <- read_csv(file_path) %>% clean_names()
  return(df)
}
```

## School Enrollment

School enrollment is a good measure to analyze how students are engaged in educational
programs and how they are thinking about their futures as specialized workers.

As children spend more time studying, they could improve their knowledge and be
able to work on a higher-demand job in the future.

In this analysis, we try to expose the influence of some indicators on students
enrollment, focused primarily on Brazil's context. The indicators that we will
cover are:

* GDP invested in education
* Children from 7 to 14 years old that are employed
* Income distribution between classes of the population

To get the data about students enrollment and the indicators, a dataset form 
World Bank Group was used and more details about it could be obtained from its
[website](https://datacatalog.worldbank.org/search/dataset/0037712/World-Development-Indicators).

We did some cleaning on the dataset so it could be more useful for our analysis.

```{r Preparing the data, message=FALSE}
# Data obtained from https://datacatalog.worldbank.org/search/dataset/0037712/World-Development-Indicators at 04/07/2022
df_countries <- read_dataset('WDICountry.csv')
df_cleaned <- read_dataset('WDIData.csv') %>%
  # Getting data until 2019 to ignore pandemic situation
  select(c(country_code, indicator_code, x2010:x2019)) %>%
  # Removing rows where all year values are null
  filter(rowSums(is.na(.)) < 10) %>%
  # Getting income groups of each country
  merge(read_dataset('WDICountry.csv'), by='country_code') %>%
  select(c(country_code, income_group, indicator_code, x2010:x2019)) %>%
  # Removing countries not associated with any Income Group
  filter(!is.na(income_group)) %>%
  # Creating a different group to Brazil, to make it comparable with other groups
  mutate(income_group=if_else(country_code == 'BRA', 'Brazil', income_group))
```

Below there is a preview of how the dataset is organized.

```{r Showing the cleaned dataset, echo=FALSE}
head(df_cleaned)
```

### How is school enrollment improving over the last years?

To get a perspective of school enrollment, we checked indicators related to
enrollments on main levels of education: primary, secondary and tertiary. The
enrollments were grouped by the Income Group of the countries as defined on
*WDICountry.csv* file. Only Brazil was taken off of its group 
(Upper middle income) to make it easier to compare with other groups.

The choice of indicators was made by using a spreadsheet with the *WDISeries.csv*
file.

```{r Definining enrollment indicators}
## Primary and secondary indicators were net indicators so they could reflect better the students that needed to be at each level
# SE.PRM.NENR: School enrollment, primary (% net)
# SE.SEC.NENR: School enrollment, secondary (% net)
# SE.TER.ENRR: School enrollment, tertiary (% gross)
enrollment_indicators <- c('SE.PRM.NENR', 'SE.SEC.NENR', 'SE.TER.ENRR')
```

After determining the indicators, we tried to group the data in a way that we 
could show the differences between groups.

```{r Grouping enrollment data}
df_enrollment <- df_cleaned %>%
  filter(indicator_code %in% enrollment_indicators) %>%
  # Getting only the last 5 year to get results based on previous years investments
  select(c(country_code, income_group, indicator_code, x2015:x2019)) %>%
  # pivoting table to make it easier to plot
  pivot_longer(x2015:x2019, names_to='year', values_to='enrollment') %>%
  # cleaning year values to make it easier to use as label
  mutate(year=gsub('x', '', .$year))
```

We also checked the presence of missing values on the metric variable of the 
dataset.

```{r Checking missing values in enrollments, echo=FALSE}
cat(
  paste(
    'Percentage of missing values on enrollment column: ',
    trunc(sum(is.na(df_enrollment$enrollment)) / nrow(df_enrollment)*100), 
    '%', 
    sep=''
    )
  )
```

And as you can see, about 40% of our enrollment variable is filled with missing 
values.As this type of value does not change much between years, I used a 
strategy to fill the missing values based on previous years, and if there isn't 
a previous year fill them with future data.

```{r Filling missing values in enrollment}
df_enrollment <- df_enrollment %>%
  group_by(country_code, indicator_code) %>%
  # Using the filling strategy to get values from previous years, if it doesn't exist, get from future year
  fill(enrollment, .direction='downup') %>%
  # Filtering to get only the countries with at least 1 not missing value for each indicator
  filter(!is.na(enrollment))
```

To check how each group is improving on their enrollment level, we plotted a 
line chart showing the differences between them.

```{r Students enrollment chart, echo=FALSE, message=FALSE, warning=FALSE}
# setting representative labels to the indicators
indicators_labels <- c(
  'SE.PRM.NENR' = 'Primary Education',
  'SE.SEC.NENR' =  'Secondary Education',
  'SE.TER.ENRR' = 'Tertiary Education'
)

df_enrollment %>%
  group_by(income_group, indicator_code, year) %>%
  summarise(mean=mean(enrollment), sd=sd(enrollment), .groups='drop') %>%
  ggplot(aes(x=year, y=mean, group=income_group)) +
  theme_minimal() +
  geom_line(aes(color=income_group), size=1) +
  geom_ribbon(aes(ymin=mean - sd, ymax=mean + sd, fill=income_group), alpha=.15) +
  expand_limits(y = 0) +
  # changing the order of the legend values
  scale_fill_brewer(breaks=legend_order, palette=color_scale) +
  scale_colour_brewer(breaks=legend_order, palette=color_scale) +
  # formatting the y labels to include a percent sign
  scale_y_continuous(labels=function(y) paste(y, '%', sep='')) +
  facet_wrap(vars(indicator_code), labeller=as_labeller(indicators_labels)) +
  labs(title='Average enrollment per Income Group and Education level', 
       subtitle='The enrollment is calculated based on the age group of the specific education level',
       x='Year', y='Enrollment', color='Income Group', fill='Income Group')
```

As we can see, there is a small difference in the Primary education level and it
increases at the other levels. It could represent a less given importance to the
other levels or some economic/social factor influences these variations. Another
interesting observation is that there is a huge standard deviation (represented
by the color shadows) inside each Income Group.

The problem with this is that children will become less capable of performing on
more specialized jobs, making it difficult to improve the overall education
quality of the countries.

In this analysis, we will give special attention to secondary education as it
can serve as a baseline to help children acquire great improvements on their
education.

```{r Secondary education enrollment, echo=FALSE, message=FALSE, warning=FALSE}
# Checking the distribution of values about secondary education enrolling for the Upper middle income countries
brazil_value <- df_enrollment %>%
  filter(income_group == 'Brazil' & indicator_code == 'SE.SEC.NENR' & year == '2019')
brazil_value <- brazil_value$enrollment

df_enrollment %>%
  filter(income_group %in% c('Brazil', 'Upper middle income') & 
           indicator_code == 'SE.SEC.NENR' & year == '2019') %>%
  ggplot(aes(x=enrollment)) + 
  theme_minimal() +
  geom_histogram(binwidth=5, fill='#80b1d3') +
  geom_vline(aes(xintercept=brazil_value), color='#757575', size=1, linetype='dashed') +
  geom_text(aes(x=brazil_value+2, y=-.3, label='Brazil')) +
  scale_x_continuous(labels=function(x) paste(x, '%', sep='')) +
  scale_y_continuous(breaks = pretty_breaks()) +
  labs(title='Enrollment in the Secondary education of the Upper middle income group on 2019', 
       x='Enrollment', y='Number of Countries')
```

According to the values about secondary school enrollment using Upper middle
income group data we could see that Brazil's enrollment level is still below a
large group of countries in the same group. So with this in mind, we check that
there is a lot of room for improvement.

### What influence does money spent on education have?

To get started on which aspects could be improved to get a better school
enrollment level, we analyzed data about GDP spent on education.

To clarify the concept, GDP 
(according to (OECD)[https://data.oecd.org/gdp/gross-domestic-product-gdp.htm]) 
"is the standard measure of the value added created through the production of 
goods and services in a country during a certain period". So we thought that it
could be a good starting point for our analysis.

The first step, as always was to group and clean the data about this indicator,
that represents the percentage of the country's GDP spent on education.

```{r Grouping GDP spent in education data}
# Checking the investment in education based on GDP
df_gdp_spent_percent <- df_cleaned %>%
  filter(indicator_code  == 'SE.XPD.TOTL.GD.ZS') %>%
  # Pivoting table to make it easier to plot
  pivot_longer(x2010:x2019, names_to='year', values_to='gdp_percent') %>%
  # Cleaning year values to make it easier to use as label
  mutate(year=gsub('x', '', .$year))
```

We checked the number of missing values to all indicators as it is a constant 
difficulty using this dataset, and in this case, we used the same strategy as the
previous indicator, to fill the missing data using previous year data and if it
doesn't exist, getting data from future years.

```{r Checking missing values in GDP spent, echo=FALSE}
cat(
  paste(
    'Percentage of missing values on gdp_percent column: ',
    trunc(sum(is.na(df_gdp_spent_percent$gdp_percent)) / nrow(df_gdp_spent_percent)*100), 
    '%', 
    sep=''
  )
)
```

```{r Filling missing values in GDP spent}
df_gdp_spent_percent <- df_gdp_spent_percent %>%
  group_by(country_code, indicator_code) %>%
  fill(gdp_percent, .direction='downup') %>%
  # filtering to get only the countries with at least 1 not missing value for each indicator
  filter(!is.na(gdp_percent))
```

```{r Percentage of GDP spent per group, echo=FALSE, message=FALSE, warning=FALSE}
df_gdp_spent_percent %>%
  group_by(income_group, indicator_code, year) %>%
  summarise(mean=mean(gdp_percent), sd=sd(gdp_percent), .groups='drop') %>%
  ggplot(aes(x=year, y=mean, group=income_group)) +
  theme_minimal() +
  geom_line(aes(color=income_group), size=1) +
  geom_ribbon(aes(ymin=mean - sd, ymax=mean + sd, fill=income_group), alpha=.15) +
  expand_limits(y = 0) +
  # Changing the order of the legend values
  scale_fill_brewer(breaks=legend_order, palette=color_scale) +
  scale_colour_brewer(breaks=legend_order, palette=color_scale) +
  # Formatting the y labels to include a percent sign
  scale_y_continuous(labels=function(y) paste(y, '%', sep='')) +
  labs(title='Average GDP spent in education', 
       x='Year', y='GDP Spent', color='Income Group', fill='Income Group')
```

The chart above shows that there is little variation in the percentage of GDP
spent on education among all countries during the last 10 years. So, we conclude
that the GDP spent on education is not relevant to our analysis.

But wait! This is only part of the analysis on GDP because it covers only the
variation in percentage and countries could have a huge difference in absolute
values of GDP and GDP per capita. So we tried to extract this type of information
and combine it with the data about expenditures on in education.

```{r Grouping GDP per capita spent in education data}
# Getting GDP per capita changes over time
df_gdp <- df_cleaned %>%
  filter(indicator_code  == 'NY.GDP.PCAP.KD') %>%
  # Pivoting table to make it easier to plot
  pivot_longer(x2010:x2019, names_to='year', values_to='gdp_per_capita') %>%
  # Cleaning year values to make it easier to use as label
  mutate(year=gsub('x', '', .$year)) %>%
  group_by(country_code, indicator_code) %>%
  # Using the strategy to fill missing values to get similar number of coutries' values
  fill(gdp_per_capita, .direction='downup') %>%
  # Filtering to get only the countries with at least 1 not missing value for each indicator
  filter(!is.na(gdp_per_capita))

df_gdp_spent_per_capita <- df_gdp_spent_percent %>%
  # Joining with gdp data to be able to get the gpd expenditure per capita
  merge(df_gdp, by=c('country_code', 'year')) %>%
  rename(income_group=income_group.x) %>%
  select(c(country_code, year, income_group, gdp_percent, gdp_per_capita)) %>%
  # Getting the gdp spent per capita
  mutate(
    gdp_spent_per_capita=(gdp_percent/100) * gdp_per_capita, 
    year=gsub('x', '', year))
```

```{r GDP spent per capita per group, echo=FALSE, message=FALSE, warning=FALSE}
df_gdp_spent_per_capita %>%
  group_by(income_group, year) %>%
  summarise(mean=mean(gdp_spent_per_capita), sd=sd(gdp_spent_per_capita), .groups='drop') %>%
  ggplot(aes(x=year, y=mean, group=income_group)) +
  theme_minimal() +
  geom_line(aes(color=income_group), size=1) +
  geom_ribbon(aes(ymin=mean - sd, ymax=mean + sd, fill=income_group), alpha=.15) +
  expand_limits(y = 0) +
  # changing the order of the legend values
  scale_fill_brewer(breaks=legend_order, palette=color_scale) +
  scale_colour_brewer(breaks=legend_order, palette=color_scale) +
  # formatting the y labels to include a percent sign
  scale_y_continuous(labels=function(y) paste('$', y, sep='')) +
  labs(title='Average GDP per capita spent in education', 
       caption='The values are based on 2015 US$',
       x='Year', y='GDP per capita spent', color='Income Group', fill='Income Group')
```

Now we can see that the difference in absolute values of GDP per capita is very
big between the High income group and other groups, with low variations during 
the last years.

```{r Distribution of GDP spent per capita of the High income group, echo=FALSE, message=FALSE, warning=FALSE}
# Let's compare how far we are from the majority of high income countries
brazil_value_gdp_per_capita <- df_gdp_spent_per_capita %>%
  filter(income_group == 'Brazil' & year == '2019')
brazil_value_gdp_per_capita <- brazil_value_gdp_per_capita$gdp_spent_per_capita

df_gdp_spent_per_capita %>%
  filter(income_group %in% c('Brazil', 'High income') & year == '2019') %>%
  ggplot(aes(x=gdp_spent_per_capita)) + 
  theme_minimal() +
  geom_histogram(binwidth=200, fill='#80b1d3') +
  geom_vline(aes(xintercept=brazil_value), color='#757575', size=1, linetype='dashed') +
  geom_text(aes(x=brazil_value+200, y=-.3, label='Brazil')) +
  scale_x_continuous(labels=function(x) paste('$', x, sep='')) +
  scale_y_continuous(breaks = pretty_breaks()) +
  labs(title='GDP per capita spent in education of the High income group on 2019 and Brazil', 
       caption='The values are based on 2015 US$',
       x='GDP per capita spent', y='Number of Countries')
```

Next, we checked how Brazil's GDP per capita is compared to the High income
group. And as we see its' value is below all other countries. It shows that we
need to increase our investments in education to get better results.

### Do children with jobs have more or fewer oportunities?

After that, the impacts of children's employment were studied to verify if there
is some correlation with children's enrollment in schools or not.

We expected there it would have, as children spending more time working will
have less time available to evolve their education.

```{r Filtering data about children employment}
# Getting data on children in employment 7 - 14 year
df_children_employment <- df_cleaned %>% filter(indicator_code  == 'SL.TLF.0714.ZS')
```

On this indicator, we changed the strategy of getting values from different years
to show the variations. This happened because the number of missing values on it is
so huge that was preferable to get only the last year with valid information.

To not get outdated information, we chose to pick the last available data
from the last 5 years.

```{r Checking missing values in children employment, echo=FALSE}
# Almost 85% of the data is missing
# Because of this we will use the strategy to get only the last year from the last five
cat(
  paste(
    'Percentage of missing values on years columns: ',
    trunc(sum(rowSums(is.na(df_children_employment))) / (nrow(df_children_employment)*10)*100), 
    '%', 
    sep=''
  )
)
```

```{r Grouping children employment}
df_children_employment <- df_children_employment %>%
  # Getting only the last year from the last five because there is a lot of missing values
  mutate(children_employment=do.call(coalesce, rev(across(x2015:x2019)))) %>%
  select(c(country_code, income_group, children_employment)) %>%
  filter(!is.na(children_employment))
```

```{r Showing children employment per income group, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(df_children_employment, aes(x=income_group, y=children_employment)) + 
  theme_minimal() +
  geom_boxplot(aes(fill=income_group)) +
  # Removing the legend as it is not necessary on this chart
  theme(legend.position='none') +
  # Changing the order of the legend values and setting default colors
  scale_x_discrete(limits=legend_order) +
  scale_y_continuous(labels=function(y) paste(y, '%', sep='')) +
  scale_fill_brewer(palette=color_scale) +
  labs(title='Percentage of children employment per income group', 
       caption='Only children with 7 - 14 years old were considered',
       x='Income Group', y='Children Employment')
```

As we expected, countries with lower incomes have more children employed than
those with higher incomes, showing a good correlation with the data about 
students enrollment.

On this chart we could not find significant data about High income countries, 
because of this, the part corresponding to it stayed blank. Analyzing Brazil's
position, we could check that it is in a good spot on this indicator.

### Is income distribution really a problem?

In the last step, we analyzed data about income share between population groups.

The population was divided into 5 categories representing the richest to the
poorest people.

```{r Definining poverty indicators}
# getting data about poverty !caution
# distribution of income share per group
# SI.DST.05TH.20: Income share held by highest 20%
# SI.DST.04TH.20: Income share held by fourth 20%
# SI.DST.03RD.20: Income share held by third 20%
# SI.DST.02ND.20: Income share held by second 20%
# SI.DST.FRST.20: Income share held by lowest 20%
poverty_indicators <- c('SI.DST.FRST.20', 'SI.DST.02ND.20' , 'SI.DST.03RD.20', 'SI.DST.04TH.20', 'SI.DST.05TH.20')
```

```{r Filtering data about income share}
df_poverty <- df_cleaned %>%
  filter(indicator_code %in% poverty_indicators) %>%
  # Pivoting table to make it easier to plot
  pivot_longer(x2010:x2019, names_to='year', values_to='income_share') %>%
  # Cleaning year values to make it easier to use as label
  mutate(year=gsub('x', '', year))
```

```{r Checking missing values in income share, echo=FALSE, results='hold'}
# Almost 50% of the data has missing values
# because of this we will use the strategy to fill years with previous values or next values
cat(
  paste(
    'Percentage of missing values on income_share column: ',
    trunc(sum(is.na(df_poverty$income_share)) / nrow(df_poverty)*100), 
    '%', 
    sep=''
  )
)
```

```{r Grouping income share values, warning=FALSE}
df_poverty <- df_poverty %>%
  group_by(country_code, indicator_code) %>%
  # Filling missing values so that each year could have the same number of countries
  fill(income_share, .direction='downup') %>%
  filter(!is.na(income_share))

df_poverty_grouped <- df_poverty %>%
  group_by(income_group, indicator_code, year) %>%
  summarise(mean=mean(income_share), sd=sd(income_share), .groups='drop_last') %>%
  # Getting the difference of values between years per group
  mutate(years_difference=abs(mean - lag(mean)))
```

To build a more interesting visualization, we checked if there is big variations
of information through the years.

With the histogram below we could see that the variations in the values are very
low, where a big part of the values varied less than 0.5% between years.

Using this information, we filtered the data to get only the last year 
information.

```{r Checking difference in income share values, echo=FALSE, message=FALSE, warning=FALSE}
# As the difference of values between years are so small, we are
# comparing values based only on the last year
cat(
  paste(
    'Maximum difference of values between years: ',
    max(df_poverty_grouped$years_difference, na.rm=TRUE), 
    '%', 
    sep=''
  )
)

ggplot(df_poverty_grouped, aes(x=years_difference)) +
  theme_minimal() +
  geom_histogram(binwidth=.025, fill='#80b1d3') +
  scale_x_continuous(labels=function(x) paste(x, '%', sep='')) +
  labs(title='Difference of income share between years', 
       x='Difference between years', y='Number of Values')
```

```{r Show income share variations, echo=FALSE, message=FALSE, warning=FALSE}
# Setting representative labels to the indicators
poverty_indicators_labels <- c(
  'Highest 20%',
  'Fourth 20%',
  'Third 20%',
  'Second 20%',
  'Lowest 20%'
)

df_poverty_grouped %>%
  filter(year == '2019') %>%
  ggplot(aes(
    # Changing the order of the legend values
    fill=factor(indicator_code, levels=rev(poverty_indicators)),
    y=mean, x=income_group)) + 
  theme_minimal() +
  geom_bar(position=position_dodge(), stat='identity') +
  # Inserting the standard deviation bars https://bioinformatics.stackexchange.com/questions/11222/stacked-bargraph-with-error-bars
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.3, position=position_dodge(0.9), color='#757575') +
  # Setting other colors to differentiate from previous charts
  # Got this custom pallet based on https://colors.dopely.top/palette-generator/ToL9uxh2Egw
  scale_fill_manual(values=c('#82AC85','#A9C7AC','#BFD8C4','#E2CFC9','#D1ACA5'), 
                    labels=poverty_indicators_labels) + 
  scale_x_discrete(limits=legend_order) +
  scale_y_continuous(labels=function(y) paste(y, '%', sep='')) +
  labs(title='Percentage of income share per Population Group and Income Group on 2019', 
       x='Income Group', y='Income Share', fill='Population Share Group')
```

On the above chart we could get that in Brazil, the differences in income share
between population groups are very big, bigger than in any other country analyzed.
These differences could increase the difficulties that poorer people have to
sustain their lives, making it harder for them to improve their education level.

We could also observe that High income countries have smaller variations through
groups showing that a more balanced income distribution could leave people with
better opportunities to continue their school enrollment.

## Conclusions

Using the World Bank data we could get some insights on which aspects Brazil
could improve to offer better opportunities to its population, showing that 
to maintain children going to school investments in education are crucial, as
well as sustaining low levels of children employment and reducing inequalities 
on income share through people.

## Future Analyses

As we discovered during the cleaning process of the data, there are a lot of
missing values through it and with this we needed to make some assumptions on how
the data would be on previous and future years. Regression algorithms could be
used to obtain better predictions on the behavior of the data, but we tried
to make the analysis simpler.

Another important aspect that could be noted is that World Bank could be 
considered a secondary source of the data on some indicators (and a good one) 
and some mistakes could have been done during the process of obtaining the data. 
So getting data from primary sources could direct us to bit different results.

## Final thoughts

In this analysis, we checked some indicators that we would believe that they 
could increase children's enrollment in schools, but it is the most important 
part of education? Maybe children that study only at home could get better 
results than those that go to school. If this is true our whole analysis could 
be changed :)

Another thing that we have to pay attention to when working with data is that
correlation is not the same as causation, so we need to be very careful with this
assumptions.

I leave you guys with a last question, if you would like to change the education
system in the whole world, what most unexpected thing would you change?

I end here with a last quote: "*Think differently, make lives better.*"